### 2025-05-04 paper/ecai2025/appendices/arch.tex
add more experiments, minor on discussion

\begin{figure}[t]
    \centering
    \includegraphics[width=.45\textwidth]{figs/attention.pdf} 
    \caption{Attention matrix for the world model.}\vspace{0.5cm}
    \label{apx:fig:attention}
\end{figure}

\begin{figure}[t]
    \centering
    \includegraphics[width=.45\textwidth]{figs/grads-mlp-transport.pdf}
    \caption{Gradient norm for a transformer architecture wrt. a MLP architecture.}\vspace{0.5cm}
    \label{fig:grads-mlp}
\end{figure}
As mentioned earlier, we test PPO, SHAC, \fname{}, and \fnamer{} with two different architecture---Transformer and MLP. The former is a popular positional invariant architecture for sequence modeling~\cite{Vaswani17}. While the latter is a simple feed-forward neural network.

>>>>

\begin{figure}[t]
    \centering
    \begin{subfigure}[b]{0.45\textwidth}
        \centering
        \includegraphics[width=\textwidth]{figs/attention.pdf}
        \caption{Attention matrix for the world model.}
        \label{apx:fig:attention}
    \end{subfigure}
    \hfill
    \begin{subfigure}[b]{0.45\textwidth}
        \centering
        \includegraphics[width=\textwidth]{figs/grads-mlp-transport.pdf}
        \caption{Gradient norm for a transformer architecture wrt. a MLP architecture.}
        \label{fig:grads-mlp}
    \end{subfigure}
    \caption{Network architecture visualizations: (left) attention pattern in the world model; (right) gradient norm comparison between transformer and MLP architectures.}
    \label{fig:combined-arch}
    \vspace{0.5cm}
\end{figure}
As mentioned earlier, we test PPO, SHAC, \fname{}, and \fnamer{} with two different architecture---Transformer and MLP. The former is a popular positional invariant architecture for sequence modeling~\cite{Vaswani17}. While the latter is a simple feed-forward neural network.

### 2025-05-04 paper/ecai2025/sects/discussion.tex
start handling some reviewer point: why we choose vmas and the scenarion, clearer intro
add more experiments, minor on discussion

\section{Discussion}
In the following, we discuss the results obtained in \Cref{sect:experiments} and provide some insights into the performance of \fname{} compared to SHAC and PPO trying to answer the research questions posed in \Cref{sect:introduction}. Note that \fname{} does not aim to outperform SHAC, it merely aims to perform comparably with SHAC while being applicable even in non-differentiable environment. 
\begin{center}
\RQ{1} \textbf{Can we train a neural network to approximate the gradients of a differentiable simulator?}
\end{center}

@@

\begin{center}
\RQ{2} \textbf{How does our algorithm compare to PPO and SHAC in both single-agent and multi-agent settings?}
\end{center}
The proposed algorithm appears capable of outperforming PPO across all scenarios showing also a better sample efficiency. In single-agent cases, while PPO manages to find near-good policies (e.g., in the Dispersion scenario), it often fails to complete scenarios within the same number of steps. Indeed, \fname{} achieves better results in all environments (see \Cref{tab:max-rewards}). Despite having to train multiple networks, \fname{} converges much faster to optimal policies, as gradient guidance enables quicker and more accurate convergence to the desired outcome. This pattern becomes even more evident in multi-agent scenarios. In these cases, PPO often struggles to find even marginally good policies.\ \fname{}, on the other hand, eventually succeeds in understanding the environment and subsequently improves the policy to enable cooperation---see both \Cref{tab:max-rewards} and \Cref{fig:experiments}.

\begin{center}
\RQ{3} \textbf{How does the performance of these algorithms change as the search space increases?}
\end{center}
To analyze the impact of increasing search spaces, we primarily focused on scenarios with varying numbers of agents. The results demonstrate that performance patterns differ across scenarios. In the Dispersion scenario, \fname{} shows increasing difficulty in finding optimal policies as the number of agents grows. Conversely, in Transport and Sampling scenarios, the system appears to converge more quickly with more agents. This counter intuitive behavior likely stems from the fact that, despite larger search spaces, the fundamental tasks in Sampling and Transport become relatively easier to solve with multiple agents, making policy discovery more straightforward. Even with approximated world dynamics (transition and reward functions), the guidance provided is sufficient to lead policies toward optimal behavior.

This advantage does not hold in the Dispersion scenario, where complexity increases substantially with additional agents. Nevertheless, we observe the emergence of cooperation across all scenarios, suggesting potential scalability to larger agent populations. The main bottleneck lies in the transformer architecture for the action world model, where the sequence length scales with the number of agents (sequence length = num.\ agents times num.\ train steps). With significantly large populations ($\sim$50 agents), the attention mechanism becomes computationally expensive, leading to prohibitively large memory requirements. In these scenarios, a linear attention mechanism (such as~\cite{Beltagy20}) may be employed.

The Discovery scenario, exhibiting high variance, shows no clear pattern compared to the other three scenarios.
This might be attributed to partial observability, which significantly complicates convergence and inter-agent cooperation.

>>>>

\section{Discussion}
In the following, we discuss the results obtained in \Cref{sect:experiments} and provide some insights into the performance of \fname{} compared to SHAC and PPO trying to answer the research questions posed in \Cref{sect:introduction}. Note that \fname{} does not aim to outperform SHAC, it merely aims to perform comparably with SHAC while being applicable even in non-differentiable environment. 
\begin{rqbox}
\textbf{Can we train a neural network to approximate the gradients of a differentiable simulator?}
\end{rqbox}

@@

\begin{replybox}
Yes, \fname{} approximates simulator gradients, achieving comparable performance to SHAC in differentiable settings, occasionally outperforming it despite initially higher variance.
\end{replybox}

\begin{rqbox}
\textbf{How does our algorithm compare to PPO and SHAC in both single-agent and multi-agent settings?}
\end{rqbox}

The proposed algorithm appears capable of outperforming PPO across all scenarios showing also a better sample efficiency.

\noindent\paragraph{Single-agent:} In single-agent cases, while PPO manages to find near-good policies (e.g., in the Dispersion scenario), it often fails to complete scenarios within the same number of steps. Indeed, \fname{} achieves better results in all environments (see \Cref{tab:max-rewards}). Compared to SHAC (where applicable), \fname{} performs comparably.

\noindent\paragraph{Multi-agent:} This pattern becomes even more evident in multi-agent scenarios. In these cases, PPO often struggles to find even marginally good policies. \fname{}, on the other hand, eventually succeeds in understanding the environment and subsequently improves the policy to enable cooperation---see both \Cref{tab:max-rewards} and \Cref{fig:experiments}. Its performance remains comparable to SHAC where SHAC can be applied.
Overall, despite having to train multiple networks, \fname{} converges much faster to optimal policies than PPO, as gradient guidance enables quicker and more accurate convergence to the desired outcome.

\begin{replybox}
\fname{} outperforms PPO across all scenarios, with PPO struggling especially in multi-agent settings. When compared to SHAC (where applicable), \fname{} achieves comparable performance while enabling emergent cooperative behaviors.
\end{replybox}

\begin{rqbox}
\textbf{How does the performance of these algorithms change as the search space increases?}
\end{rqbox}

To analyze the impact of increasing search spaces, we primarily focused on scenarios with varying numbers of agents (up to 5 in the main experiments). The results demonstrate that performance patterns differ across scenarios. In the Dispersion scenario, \fname{} shows increasing difficulty in finding optimal policies as the number of agents grows. Conversely, in Transport and Sampling scenarios, the system appears to converge more quickly with more agents. This counter intuitive behavior likely stems from the fact that, despite larger search spaces, the fundamental tasks in Sampling and Transport become relatively easier to solve with multiple agents, making policy discovery more straightforward. Even with approximated world dynamics (transition and reward functions), the guidance provided is sufficient to lead policies toward optimal behavior.

This scaling advantage is not observed in the Dispersion scenario, where complexity increases quadratically with agent count. Nevertheless, we observe consistent emergence of cooperative behavior patterns across all experimental scenarios, suggesting potential scalability to larger agent populations (quantitative analysis with up to 20 agents is provided in \Cref{apx:experiments}). The primary computational constraint lies in the transformer architecture employed for the action world model, where sequence length scales linearly with the number of agents (specifically, sequence length = $n \times s$, where $n$ is the number of agents and $s$ is the number of training steps). With agent populations exceeding approximately 50 agents, the self-attention mechanism's $O(n^2)$ complexity leads to prohibitive memory requirements. For such large-scale multi-agent systems, linear attention mechanisms (e.g.,~\cite{Beltagy20}) represent a viable architectural alternative.

\begin{replybox}
Performance varies with scale: Transport/Sampling converge faster with more agents, while Dispersion becomes harder. Cooperation emerges across scenarios. Main scalability limitation is transformer attention complexity, becoming prohibitive at large agent counts ($~$50), though moderate populations remain feasible.
\end{replybox}

### 2025-05-04 paper/ecai2025/sects/experiments.tex
add more experiments, minor on discussion

\begin{figure*}[t]
    \includegraphics[width=\textwidth]{figs/main-transformer.pdf}
    \caption{Performance comparison of \fname{}, PPO, and SHAC across different scenarios (Dispersion, Transport, Discovery, and Sampling) using the Transformer architecture for multi-agent settings and the MLP architecture for single-agent settings. The results show the mean and standard deviation of rewards over 3 runs. For results using the MLP architecture in all scenarios, refer to \Cref{apx:fig:experiments-mlp}.}%
    \label{fig:experiments}
\end{figure*}

\begin{figure}[t]
    \centering
    \includegraphics[width=\columnwidth]{figs/grads-transformer-transport.pdf}
    \caption{Norm of policy gradients on the Transport environment across 5000 epochs. The raise in norm for SHAC (epoch 2700) corresponds with occurrence of generalization, where collision between agents and the packages become frequent. Also \fname{} experience a raise in norm, albeit more behaved, corresponds with the occurrence of generalization (epoch 3200).}\label{fig:grads-transformer-transport}
    \vspace{0.5cm}
\end{figure}
\section{Experiments}\label{sect:experiments}

@@

\subsection{Results}
\begin{figure}
    \centering
    \includegraphics[width=0.32\columnwidth]{figs/step-1.png}
    \includegraphics[width=0.32\columnwidth]{figs/step-2.png}
    \includegraphics[width=0.32\columnwidth]{figs/step-3.png}
    \caption{Snapshots of the result of the Transport scenario. The agents are represented as colored circles, the package as a red square, and the goal as a green circle. The agents are trained to push the package to the goal.}\label{fig:transport-trained}
    \vspace{0.5cm}
\end{figure}
In this section, we compare the performance of \fname{}, PPO, and SHAC across different scenarios using the Transformer architecture for multi-agent settings and the MLP architecture for single-agent settings. The results show the mean and standard deviation of achieved rewards over 3 runs. An overview of the results is shown in \Cref{fig:experiments} and \Cref{tab:max-rewards}. For results using the MLP architecture in all scenarios, refer to \Cref{apx:fig:experiments-mlp}.

>>>>

\begin{figure*}[t]
    \centering
    \includegraphics[width=0.9\textwidth]{figs/main-transformer.pdf}
    \caption{Performance comparison of \fname{}, PPO, and SHAC across different scenarios (Dispersion, Transport, Discovery, and Sampling) using the Transformer architecture for multi-agent settings and the MLP architecture for single-agent settings. The results show the mean and standard deviation of rewards over 3 runs. For results using the MLP architecture in all scenarios, refer to \Cref{apx:fig:experiments-mlp}.}%
    \label{fig:experiments}
\end{figure*}

\begin{figure}[t]
    \centering
    \includegraphics[width=0.9\columnwidth]{figs/grads-transformer-transport.pdf}
    \caption{Policy gradient norms in the Transport environment over 5000 epochs. Spikes in gradient norm for SHAC (epoch 2700) and \fname{} (epoch 3200) correspond to generalization events when agents begin frequent interactions with the package and each other.}\label{fig:grads-transformer-transport}
    \vspace{0.5cm}
\end{figure}
\section{Experiments}\label{sect:experiments}

@@

\subsection{Results}
\begin{figure}
    \centering
    \includegraphics[width=0.32\columnwidth]{figs/step-1.png}
    \includegraphics[width=0.32\columnwidth]{figs/step-2.png}
    \includegraphics[width=0.32\columnwidth]{figs/step-3.png}
    \caption{Snapshots of trained agents with \fname{} in the Transport scenario. Colored circles represent agents pushing a red package (square) toward the green goal.}\label{fig:transport-trained}
    \vspace{0.5cm}
\end{figure}
In this section, we compare the performance of \fname{}, PPO, and SHAC across different scenarios using the Transformer architecture for multi-agent settings and the MLP architecture for single-agent settings. The results show the mean and standard deviation of achieved rewards over 3 runs. An overview of the results is shown in \Cref{fig:experiments} and \Cref{tab:max-rewards}. For results using the MLP architecture in all scenarios, refer to \Cref{apx:fig:experiments-mlp}.
